Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-01-14 doc/invt.tex
invt.tex, add the proof of optimality of WL scheme

To do so, we need to introduce a bias potential $\phi_i$,
such that the target distribution is changed to
%
\begin{equation}
  \pi_i \propto p^*_i \, e^{-\phi_i}.
  \label{eq:pi_p_phi1}
\end{equation}
%
Upon normalization $\sum_{i = 1}^n \pi_i = 1$,
we get
%
\begin{equation}
  \pi_i =
  \frac{ p^*_i \, e^{-\phi_i} }
  { \sum_{j = 1}^n p^*_j \, e^{-\phi_j} }.
  \label{eq:pi_p_phi}
\end{equation}
%
Particularly,
to achieve a flat distribution $\pi_i$,
the bias potential $\phi_i$
must coincide with $\log p^*_i$,
up to an additive constant.

Generally, we may wish to sample
a nearly flat distribution, $p_i$,
by adjusting the bias potential $\phi_i$
on the fly.
%
For example,
we can monitor the histogram accumulated over
a period of time and use it for $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1},
and then update
$\phi_i$ according to
$$
\phi_i^{\mathrm{new}}
=
\phi_i^{\mathrm{old}}
-
\log \frac{ p_i } { \pi_i }.
$$
%
%From Eq. \eqref{eq:pi_p_phi1},
%it is clear that to achieve the target distribution
%$\pi_i = p_i$ requires
%%
%\begin{equation}
%\log p^*_i
%=\phi_i + \ln p_i + c
%\label{eq:bias_pmf}
%\end{equation}
%%
%%
%In the end of the simulation,
%the PMF, $\phi^*_i$,
%can be deduced from Eq. \eqref{eq:bias_pmf}.
%
The above inversion process can be inconvenient
in practice,
for we need to make the period
as long as a multiple of
the autocorrelation time,
which is difficult to estimate for an unknown system.

@@

The Wang-Landau (WL) algorithm is a convenient technique
of obtaining the desired bias potential $\phi_i$
on the fly.
%
To simplify the notation, we introduce
a shifted bias potential
%
\begin{equation}
  v_i \equiv \phi_i - \ln p^*_i + \ln p_i.
  \label{eq:v_def}
\end{equation}
%
%
and in $v_i$, Eq. \eqref{eq:pi_p_phi}
becomes
%
\begin{equation}
  \pi_i = \frac{ p_i e^{-v_i} }
  { \sum_{j = 1}^n p_j e^{-v_j} }
  \propto p_i \, e^{-v_i}.
  \label{eq:pi_p_v}
\end{equation}
%
According to Eq. \eqref{eq:pi_p_phi1},
$v_i$ should approach a constant of $i$
upon convergence (i.e., when $\pi_i \approx p_i$).

@@

Next, we split $h_i(t)$ into a deterministic part
and a noise part,
%
\begin{equation}
  h_i(t) = \langle h_i(t) \rangle + \zeta_i(t).
  \label{eq:h_split}
\end{equation}
%
Here, the deterministic part can be related
to the ``ensemble average'' of $h_i$.
%
The ensemble consists of many similar simulation copies
that have experienced the same schedule $\alpha(t)$
and have reached the same $v_i(t)$
at time $t$.
%
The initial states and random numbers of the copies
may, however, be different.

For sufficiently small $\alpha(t)$,
the sampling process is approximately
Markovian of a finite order,
and we may assume Eq. \eqref{eq:pi_p_v}
for the deterministic part
%
\begin{equation}
  \langle h_i(t) \rangle
  \approx
  \frac{ p_i \, e^{-v_i} }
  { \sum_{j = 1}^n p_j \, e^{-v_j} }.
  \label{eq:h_ave}
\end{equation}

For the noise part, we have
$\langle \zeta_i(t) \rangle = 0$.
%
%The noise is not necessarily white.
We consider a general noise with the correlation
function depending on the time difference:
%
\begin{equation}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \sigma_{ij}(t - t'),
  \label{eq:zeta_zeta_correlation}
\end{equation}
%
where $\sigma_{ij}(t)$ is an even function of $t$
that vanishes at large $t$.

@@

Further, by Eq. \eqref{eq:w_detailedbalance},
we can define a symmetric matrix $\hat{\mathbf w}$
as $\hat w_{ij} = \sqrt{p_i/p_j} \, w_{ij}$
that can be diagonalized
with a set of orthonormal eigenvectors:
%
$\sum_{i = 1}^n U_{ki} \, \hat w_{ij} = \lambda_k \, U_{kj}$.
%
Thus,
in terms of diagonalizing $\mathbf w$, we have
\begin{equation}
  \sum_{i = 1}^n T_{ki} \, w_{ij} = \lambda_k \, T_{kj},
  \label{eq:T_w}
\end{equation}
where
$T_{ij} = \sqrt{p_j} \, U_{ij}$ satisfies
the orthonormal conditions:
%
\begin{align}
\sum_{k = 1}^n T_{ki} \, T_{kj}
&= \delta_{ij} \, p_i,
\label{eq:T_orthonormal_cols}
\\
\sum_{k = 1}^n \frac{ T_{ik} \, T_{jk} }{ p_k }
&= \delta_{ij}.
\label{eq:T_orthonormal_rows}
\end{align}

We emphasize that, in a stable updating scheme,
all eigenvalues $\lambda_k$ must be nonnegative,
for otherwise there is a fluctuating mode
that can increase indefinitely.

@@

Now by diagonalizing the matrix $\mathbf w$
using matrix defined in Eqs. \eqref{eq:T_w},
\eqref{eq:T_orthonormal_cols},
and
\eqref{eq:T_orthonormal_rows},
we get a set of decoupled equations
%
\begin{equation}
\dot y_i(t)
=
-\alpha(t) \, \lambda_i
[y_i(t) - \eta_i(t)],
\label{eq:yt_diffeq}
\end{equation}
%
where
\begin{align}
  y_i &= \sum_{j=1}^n T_{ij} \, x_j,
  \label{eq:y_def}
  \\
  \eta_i &= \sum_{j=1}^n T_{ij} \frac{ \zeta_j}{ p_j}.
  \label{eq:eta_def}
\end{align}

@@

The error function can be obtained
from Eq. \eqref{eq:T_orthonormal_cols}
as
\begin{align}
  \mathcal E
  &\equiv
  \sum_{i = 1}^n p_i \, x_i^2
  =
  \sum_{i, j, k=1}^n T_{ki} \, T_{kj} \, x_i \, x_j
  =
  \sum_{k = 1}^n y_k^2.
  \label{eq:y2_sum}
\end{align}
%
We note that there is one mode, say
$y_1 = \sum_{j=1}^n p_j \, x_j$
(corresponding to $\lambda_1 = 1$),
vanishes identically by Eq. \eqref{eq:px_sum}.
%
Thus, we can start the sum in Eq. \eqref{eq:y2_sum}
from $k = 2$.

In analogous to Eq. \eqref{eq:x2t_average},
we can rewrite Eq. \eqref{eq:y2_sum} as
%
\begin{align}
  \mathcal E
  %&= \sum_{i = 2}^n \langle y_i^2 \rangle
  %\notag
  %\\
  &=
  \int_0^t \int_0^t
  \sum_{i = 2}^n
  \dot u_i\bigl( q(t') \bigr) \,
  \dot u_i\bigl( q(t'') \bigr) \,
  \kappa_i(t' - t'') \, dt' \, dt'',
  \label{eq:error_mbin}
\end{align}
%
where
\begin{align*}
  u_i(z)
  &\equiv
  e^{\lambda_i \, [q(t') - z]},
  \\
  \kappa_i(t - t')
  &\equiv
  \left\langle
    \eta_i(t) \, \eta_i(t')
  \right\rangle,
\end{align*}
%
and the extremal condition of $\mathcal E$ is given by
%
\begin{align}
\sum_{i=2}^n
\lambda_i \, u_i\bigl( q(\tau) \bigr)
\int_0^t
\ddot u_i\bigl( q(t') \bigr) \, \kappa_i(t' - \tau) \, dt' = 0.
\label{eq:optimal_mbin}
\end{align}
%
If all $\lambda_i$, hence all $u_i$, are the same,
the above condition can be satisfied
by Eq. \eqref{eq:ddu_eq_0},
which is the single-bin case.
%
The general solution of Eq. \eqref{eq:optimal_mbin}
is harder to obtain.

To proceed, we shall further assume perfect sampling,
%
\begin{equation}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  p_i \, \delta_{ij} \, \delta(t - t').
  \label{eq:zz_perfect}
\end{equation}
%
Thus, from Eqs. \eqref{eq:T_orthonormal_rows} and \eqref{eq:eta_def},
we get
%
\begin{equation}
  \kappa_i(t - t')
  =
  \sum_{j,k = 1}^n
  \frac{ T_{ij} } { p_j }
  \frac{ T_{ik} } { p_k }
  \langle \zeta_j(t) \, \zeta_k(t') \rangle
  =
  \delta(t - t').
  \label{eq:kappa_perfect}
\end{equation}
%
Then, Eq. \eqref{eq:error_mbin} is simplified as
%
\begin{align}
  \mathcal E
  &=
  \int_0^t
  \sum_{i = 2}^n
  \dot u_i^2\bigl( q(t') \bigr) \, dt',
  \label{eq:error_mbin_perfect}
\end{align}
%
and Eq. \eqref{eq:optimal_mbin} becomes
%
\begin{align}
  \sum_{i=1}^n
  \lambda_i \, u_i\bigl( q(\tau) \bigr) \, \ddot u_i\bigl( q(\tau) \bigr) = 0.
  \label{eq:optimal_mbin_perfect}
\end{align}
%
\note{If we treat $u_i$ as the position,
  and $\lambda_i$ as the mass,
  then $\lambda_i \, \ddot u_i$ is the force.
  So the left-hand side of Eq. \eqref{eq:optimal_mbin_perfect}
  is the virial.
  %
  Thus this condition states that the total virial is zero.
}

@@

@@

A more direct, and often more accurate, approach
is to assume the functional form
%
\begin{equation}
\alpha(\tau) = \frac{1}{\lambda \, (\tau + t_0) },
\label{eq:alpha_trial}
\end{equation}
%
and to find the optimal value of $\lambda$.
%
Using Eq. \eqref{eq:alpha_trial}
in Eq. \eqref{eq:error_mbin_perfect} yields
%
\begin{align}
\mathcal E
&=
\frac{1}{t+t_0}
\sum_{i=2}^n \frac{ \lambda_i^2 } { \lambda \, (2 \lambda_i - \lambda) }
\left[
  1 - \left(
    1 + \frac{t}{t_0}
  \right)^{1 - 2 \, \lambda_i/\lambda}
\right]
\notag
\\
&\stackrel{      t \gg t_0      }
          { =\joinrel=\joinrel= }
%
\sum_{i = 2}^n
\begin{dcases}
  \frac{ 1 }
       { t }
  \frac{ \lambda_i^2 }
       { \lambda \, (2 \, \lambda_i - \lambda) }
  &
  \mathrm{if \;} \lambda < 2 \, \lambda_i,
  \\%[1em]
  %
  %
  \frac{   1    }
       { 4 \, t }
  \ln \frac{ t + t_0 }
           {  t_0    }
  &
  \mathrm{if \;} \lambda = 2 \, \lambda_i,
  \\%[1em]
  %
  %
  \frac{ t_0^{ 2 \, \lambda_i / \lambda  - 1} }
       { t^{ 2 \, \lambda_i / \lambda } }
  \frac{ \lambda_i^2 }
       { \lambda \, (\lambda - 2 \, \lambda_i) }
  &
  \mathrm{if \;} \lambda > 2 \, \lambda_i.
\end{dcases}
\label{eq:y2_sum_perfect_invt}
\end{align}
%
\note{In the above expression,
$t + t_0$ are replaced by $t$.
%
Adding back $t_0$, we get
$$
\begin{aligned}
  \mathcal E
  \stackrel{         t \to \infty         }
           { =\joinrel=\joinrel=\joinrel= }
  \sum_{i = 2}^n
  \begin{dcases}
    \frac{ 1 }{ t + t_0 }
    \frac{ \lambda_i^2 } { \lambda \, (2 \, \lambda_i - \lambda) }
    &
    \mathrm{if \;} \lambda < 2 \, \lambda_i,
    \\%[1em]
    %
    %
    \frac{1}{4 \, (t + t_0) } \ln\frac{t + t_0}{t_0}
    &
    \mathrm{if \;} \lambda = 2 \, \lambda_i,
    \\
    %
    %
    \frac{ t_0^{ 2 \, \lambda_i / \lambda  - 1} }
         { (t + t_0)^{ 2 \, \lambda_i / \lambda } }
    \frac{ \lambda_i^2 }
         { \lambda \, (\lambda - 2 \, \lambda_i) }
    &
    \mathrm{if \;} \lambda > 2 \, \lambda_i.
  \end{dcases}
\end{aligned}
$$
}
%
The last two cases, $\lambda \ge 2\lambda_i$,
are asymptotically suboptimal
as they decay slower than $1/t$.
%
Thus, the optimal $\lambda$
satisfies $\lambda < 2 \min \lambda_i$,
and in this case,
we can seek that
the optimal value from
the smallest real root of
%
\begin{equation}
\sum_{i = 2}^n
\frac{ \lambda_i - \lambda }
{ \left(2 - \lambda/ \lambda_i \right)^2 }
= 0.
\label{eq:optimal_lambda_approx}
\end{equation}

Under the approximation, we have
$$
\frac{ \lambda_i^2 }{ \lambda \, (2 \lambda_i - \lambda) } \ge 1,
$$
with the equality achieved only at $\lambda = \lambda_i$,
we see the from Eq. \eqref{eq:y2_sum_perfect_invt}
that the minimal error is achieved with
\begin{equation}
  \lambda_2 = \dots = \lambda_n = \lambda.
\end{equation}
%
and the updating matrix $\mathbf w$ satisfying this condition
is essentially a multiple of the identity matrix
(since $\lambda_1$ corresponds to the mode
of a uniform shift of all $v_i$,
we can freely set it to $\lambda_2$
without changing the nature of the updating scheme).
%
This shows that in terms of convergence,
the single-bin updating scheme
(as adopted in the WL algorithm)
is the most efficient candidate
of all multiple-bin schemes.

@@

As an example, consider the following matrix $\mathbf w$
%
\begin{equation}
  w_{ij} = m_{|i-j|} + m_{i+j-1} + m_{2n+1-i-j},
  \label{eq:w_band}
\end{equation}
%
where $m_0, \dots, m_b \, (b < n)$
are a set of numbers satisfying
%
\begin{equation}
m_0 + 2 \, m_1 + \cdots + 2 \, m_b = 1,
\label{eq:m_normalization}
\end{equation}
%
and $m_l = 0$ for $l > b$.

This matrix, being symmetric,
satisfies Eqs. \eqref{eq:w_sumj}-\eqref{eq:w_balance},
for the flat distribution $p_i \equiv 1/n$.
%
It admits $n$ eigenvectors,
$\pmb\theta^{(1)}, \dots, \pmb\theta^{(n)}$,
given by
%
\begin{equation}
\theta^{(k)}_i
= \cos \frac{ \left(i - \frac12\right) (k - 1) \, \pi}{n}.
\label{eq:wband_eigenvector}
\end{equation}
%
with eigenvalues
\begin{equation}
  \lambda_k = m_0 + \sum_{l = 1}^b 2 \, m_l \cos \frac{(k - 1)  \, l \pi}{n}.
\label{eq:wband_eigenvalue}
\end{equation}

>>>>

To do so, we need to introduce a bias potential $V_i$,
such that the target distribution is changed to
%
\begin{equation}
  \pi_i \propto p^*_i \, e^{-V_i}.
  \label{eq:pi_p_phi1}
\end{equation}
%
Upon normalization $\sum_{i = 1}^n \pi_i = 1$,
we get
%
\begin{equation}
  \pi_i =
  \frac{ p^*_i \, e^{-V_i} }
  { \sum_{j = 1}^n p^*_j \, e^{-V_j} }.
  \label{eq:pi_p_phi}
\end{equation}
%
Particularly,
to achieve a flat distribution $\pi_i$,
the bias potential $V_i$
must coincide with $\log p^*_i$,
up to an additive constant.

Generally, we may wish to sample
a nearly flat distribution, $p_i$,
by adjusting the bias potential $V_i$
on the fly.
%
For example,
we can monitor the histogram accumulated over
a period of time and use it for $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1},
and then update
$V_i$ according to
$$
V_i^{\mathrm{new}}
=
V_i^{\mathrm{old}}
-
\log \frac{ p_i } { \pi_i }.
$$
%
%From Eq. \eqref{eq:pi_p_phi1},
%it is clear that to achieve the target distribution
%$\pi_i = p_i$ requires
%%
%\begin{equation}
%\log p^*_i
%=V_i + \ln p_i + c
%\label{eq:bias_pmf}
%\end{equation}
%%
%%
%In the end of the simulation,
%the PMF, $-\log p^*_i$,
%can be deduced from Eq. \eqref{eq:bias_pmf}.
%
The above inversion process,
adopted by early multicanonical or entropic sampling,
can be inconvenient in practice,
for we need to make the period
as long as a multiple of
the autocorrelation time,
which is difficult to estimate for an unknown system.

@@

The Wang-Landau (WL) algorithm is a convenient technique
of obtaining the desired bias potential $V_i$
on the fly.
%
To simplify the notation, we introduce
a shifted bias potential
%
\begin{equation}
  v_i \equiv V_i - \ln p^*_i + \ln p_i.
  \label{eq:v_def}
\end{equation}
%
%
and in $v_i$, Eq. \eqref{eq:pi_p_phi}
becomes
%
\begin{equation}
  \pi_i
  =
  \frac{                p_i \, e^{-v_i} }
       { \sum_{j = 1}^n p_j \, e^{-v_j} }
  \propto
  p_i \, e^{-v_i}.
  \label{eq:pi_p_v}
\end{equation}
%
According to Eq. \eqref{eq:pi_p_phi1},
$v_i$ should approach a constant of $i$
upon convergence (i.e., when $\pi_i \approx p_i$).

@@

Next, we split $h_i(t)$ into a deterministic part
and a noise part,
%
\begin{equation}
  h_i(t) = \langle h_i(t) \rangle + \zeta_i(t).
  \label{eq:h_split}
\end{equation}
%
Here, the deterministic part can be related
to the ``ensemble average'' of $h_i$.
%
The ensemble consists of many similar simulation copies
that have experienced the same schedule $\alpha(t)$
and have reached the same $v_i(t)$
at time $t$.
%
The initial states and the stochastic forces
during the process may, however, be different.

For sufficiently small $\alpha(t)$,
the $v_i$'s remain roughly the same for a short period.
%
Then,
the sampling process is approximately Markovian, % of a finite order,
and we may assume Eq. \eqref{eq:pi_p_v}
for the deterministic part
%
\begin{equation}
  \langle h_i(t) \rangle
  \approx
  \pi_i
  =
  \frac{ p_i \, e^{-v_i} }
  { \sum_{j = 1}^n p_j \, e^{-v_j} }.
  \label{eq:h_ave}
\end{equation}
%
%
%
For the noise part, we have
$\langle \zeta_i(t) \rangle = 0$,
%
%The noise is not necessarily white.
and the correlation functions
depend only on the time difference
%
\begin{equation}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \sigma_{ij}(t - t'),
  \label{eq:zeta_zeta_correlation}
\end{equation}
%
where $\sigma_{ij}(t)$ is an even function of $t$
that vanishes at large $t$.
%
More explicitly,
if the transition matrix of a period $\tau$
is $T^\tau_{ij}$,
then
$
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \pi_j - \pi_i \, \pi_j.
$

@@

Further, by Eq. \eqref{eq:w_detailedbalance},
we can define a symmetric matrix $\hat{\mathbf w}$
as $\hat w_{ij} = \sqrt{p_i/p_j} \, w_{ij}$
that can be diagonalized
with a set of orthonormal eigenvectors:
%
$\sum_{i = 1}^n \varphi_{ki} \, \hat w_{ij} = \lambda_k \, \varphi_{kj}$.
%
Thus,
in terms of diagonalizing $\mathbf w$, we have
\begin{equation}
  \sum_{i = 1}^n \phi_{ki} \, w_{ij} = \lambda_k \, \phi_{kj},
  \label{eq:eig_w}
\end{equation}
where
$\phi_{ij} = \sqrt{p_j} \, \varphi_{ij}$ satisfies
the orthonormal conditions:
%
\begin{align}
\sum_{k = 1}^n \phi_{ki} \, \phi_{kj}
&= \delta_{ij} \, p_i,
\label{eq:eig_orthonormal_cols}
\\
\sum_{k = 1}^n \frac{ \phi_{ik} \, \phi_{jk} }{ p_k }
&= \delta_{ij}.
\label{eq:eig_orthonormal_rows}
\end{align}

%It is worth pointing out
Note that, in a stable updating scheme,
all eigenvalues $\lambda_k$ must be nonnegative,
for otherwise there is a fluctuating mode
that can increase indefinitely.

@@

Now by diagonalizing the matrix $\mathbf w$
using matrix defined in Eqs. \eqref{eq:eig_w},
\eqref{eq:eig_orthonormal_cols},
and
\eqref{eq:eig_orthonormal_rows},
we get a set of decoupled equations
%
\begin{equation}
\dot y_i(t)
=
-\alpha(t) \, \lambda_i
[y_i(t) - \eta_i(t)],
\label{eq:yt_diffeq}
\end{equation}
%
where
\begin{align}
  y_i &= \sum_{j=1}^n \phi_{ij} \, x_j,
  \label{eq:y_def}
  \\
  \eta_i &= \sum_{j=1}^n \phi_{ij} \frac{ \zeta_j}{ p_j}.
  \label{eq:eta_def}
\end{align}

@@

The error function can be obtained
from Eq. \eqref{eq:eig_orthonormal_cols}
as
\begin{align}
  \mathcal E
  &\equiv
  \sum_{i = 1}^n p_i \, x_i^2
  =
  \sum_{i, j, k=1}^n \phi_{ki} \, \phi_{kj} \, x_i \, x_j
  =
  \sum_{k = 1}^n y_k^2.
  \label{eq:y2_sum}
\end{align}
%
We note that there is one mode, say
$y_1 = \sum_{j=1}^n p_j \, x_j$
(corresponding to $\lambda_1 = 1$),
vanishes identically by Eq. \eqref{eq:px_sum}.
%
Thus, we can start the sum in Eq. \eqref{eq:y2_sum}
from $k = 2$.

In analogous to Eq. \eqref{eq:x2t_average},
we can rewrite Eq. \eqref{eq:y2_sum} as
%
\begin{align}
  \mathcal E
  %&= \sum_{i = 2}^n \langle y_i^2 \rangle
  %\notag
  %\\
  &=
  \int_0^t \int_0^t
  \sum_{i = 2}^n
  \dot u_i\bigl( q(t') \bigr) \,
  \dot u_i\bigl( q(t'') \bigr) \,
  \kappa_i(t' - t'') \, dt' \, dt'',
  \label{eq:error_mbin}
\end{align}
%
where
\begin{align*}
  u_i(z)
  &\equiv
  e^{\lambda_i \, [q(t') - z]},
  \\
  \kappa_i(t - t')
  &\equiv
  \left\langle
    \eta_i(t) \, \eta_i(t')
  \right\rangle.
\end{align*}
%
\note{The extremal condition of $\mathcal E$ is given by
%
\begin{align}
\sum_{i=2}^n
\lambda_i \, u_i\bigl( q(\tau) \bigr)
\int_0^t
\ddot u_i\bigl( q(t') \bigr) \, \kappa_i(t' - \tau) \, dt' = 0.
\label{eq:optimal_mbin}
\end{align}
%
If all $\lambda_i$, hence all $u_i$, are the same,
the above condition can be satisfied
by Eq. \eqref{eq:ddu_eq_0},
which is the single-bin case.
%
Unfortunately, the above expression
appears to be useless in a general setting.
}

\subsection{White-noise approximation}

In the asymptotically regime,
where $t \gg 1/\alpha(t)$,
we can approximate $\kappa_i(t)$ as
%
\begin{equation}
  \kappa_i(t) \approx \Gamma_i \, \delta(t)
  \label{eq:kappa_delta}
\end{equation}
%
where, $\Gamma_i = \int_{-\infty}^\infty \kappa_i(t) \, dt$.
%
Then, Eq. \eqref{eq:error_mbin} is simplified as
%
\begin{align}
  \mathcal E
  &=
  \int_0^t
  \sum_{i = 2}^n
  \Gamma_i \, \dot u_i^2\bigl( q(t') \bigr) \, dt',
  \label{eq:error_mbin1}
\end{align}
%
%
\note{This is equivalent to saying that the Fourier transform
  $$
  \tilde \kappa_i(\omega) = \Gamma_i
  $$
  is a roughly constant.
  %
  For the next order correction we would have
  $$
  \tilde \kappa_i(\omega) = \Gamma_i + \Gamma^{(2)}_i \omega^2 + \dots,
  $$
  which leads to a correction to $\mathcal E$
  $$
  \Gamma^{(2)}_i
  \int_0^t \ddot u_i^2\bigl( q(t') \bigr) \, dt'.
  $$
  This correction is of order $\alpha^2$
  compared to the main term in Eq. \eqref{eq:error_mbin1},
  and hence asymptotically unimportant.

  The extremal property, Eq. \eqref{eq:optimal_mbin},
  now becomes
  %
  \begin{align}
    \sum_{i=1}^n
      \Gamma_i \, \lambda_i \,
      u_i\bigl( q(\tau) \bigr) \,
      \ddot u_i\bigl( q(\tau) \bigr) = 0.
    \label{eq:optimal_mbin1}
  \end{align}
  %
  If we interpret $u_i$ as the position,
  and $\Gamma_i \, \lambda_i$ as the mass,
  then the right-hand side give the virial.
  %
  Thus, the extremal condition demands
  the total virial to be zero.
  $$\,$$
}
%
For example,
if we assume perfect sampling,
%
\begin{equation}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  p_i \, (1 - p_i) \, \delta_{ij} \, \delta(t - t').
  \label{eq:zz_perfect}
\end{equation}
%
Thus, from Eqs. \eqref{eq:eig_orthonormal_rows} and \eqref{eq:eta_def},
we get
%
\begin{equation}
  \kappa_i(t - t')
  =
  \sum_{j,k = 1}^n
  \frac{ \phi_{ij} } { p_j }
  \frac{ \phi_{ik} } { p_k }
  \langle \zeta_j(t) \, \zeta_k(t') \rangle
  %
  \approx \delta(t - t').
  \label{eq:kappa_perfect}
\end{equation}
%
where we have assumed the large $n$ limit
such that $p_i \ll 1$
in the last expression.

@@

\subsection{\label{sec:optWL}
Optimality of the single-bin updating scheme}

We can show from Eq. \eqref{eq:error_mbin1}
that the single-bin updating scheme
is an optimal updating scheme.
%
Using the Cauchy-Schwarz inequality, we have,
for any set of $\beta_i \ge 0$
%
\begin{align}
&
\left( \int_0^t
  dt' \sum_{i = 2}^n \Gamma_i \,
                     \dot u_i^2\bigl( q(t') \bigr)
\right)
\left(
\int_0^t
  dt' \sum_{i = 2}^n \Gamma_i \, c_i^2
\right)
\notag
\\
&
\ge
\left(
\int_0^t
  dt' \sum_{i = 2}^n \Gamma_i \, c_i \,
                     \dot u_i\bigl( q(t') \bigr)
\right)^2
\notag
\\
&
=
\left(
  \sum_{i = 2}^n \Gamma_k \, c_i
    \left[
      1 - e^{ -\lambda_i q(t) }
    \right]
\right)^2.
\notag
\end{align}
%
\note{A quick proof of the above version
  of the Cauchy-Schwarz inequality
  is the following.
  %
  First, we have
$$
\int_0^t
  dt' \sum_{i = 2}^n \Gamma_i \,
    \left( \dot u_i \, X - c_i \right)^2
\ge 0,
$$
  for any $X$.
  %
  Thus, the right-hand side,
  as a quadratic polynomial of $X$,
  must have a non-positive discriminant.
  %
  This condition gives the desired inequality.
  \\%
}
%
Note that the last expression depends on
$q(t')$ only through the end point value,
which is fixed in the variation process.
%
The equality is achieved
if $\dot u_i\left( q(t') \right) = c_i$
for all $i = 2, \dots, n$ at any $t'$,
up to a multiplicative factor.
%
Solving this set of equations yields
$$
\alpha(t') = \frac{              1              }
                  { \lambda_i \, (t' + t_{i0} ) },
\quad
\mathrm{with\;\;}
t_{i0} = \frac{             t            }
              { e^{ \lambda_i q(t) } - 1 }.
$$
Such a solution is possible only if
\begin{equation}
\lambda_2 = \dots = \lambda_n.
\end{equation}
%
\note{
  We add some steps for solution.
  Integrating $\dot u_i = c_i$ yields
  \begin{equation}
  u_i(t') = c_i \left(t' + t_{i0} \right).
  \label{eq:ui_solution}
  \end{equation}
  Given that $u_i(t) = 1$ and $u_i(0) = e^{-\lambda_i \, q(t)}$,
  we have
  $$
  c_i = \frac{ 1 }{ t + t_{i0} },
  \quad
  \mathrm{and\;}
  \frac{ t_{i0} } { t + t_{i0} }
  =
  e^{ -\lambda_k \, q(t) }.
  $$
  Taking the logarithm of Eq. \eqref{eq:ui_solution} yields
  $$
  -\lambda_i \, q(t) + \lambda_i \, q(t')
  = \log c_i + \log\left( t' + t_{i0} \right).
  $$
  Differentiating this with respect to $t'$ yields
  $$
  \lambda_i \, \alpha(t') = \frac{ 1 } { t' + t_{i0} }.
  $$
}

Now an updating matrix $\mathbf w$ satisfying this condition
is essentially a multiple of the identity matrix
(since $\lambda_1$ corresponds to the mode
of a uniform shift of all $v_i$,
we can freely set it to $\lambda_2$
without changing the nature of the updating scheme).
%
Thus, in terms of convergence,
the single-bin updating scheme
(as adopted in the WL algorithm)
is generally the most efficient
updating scheme.

@@

If we wish to find,
for a given updating matrix $\mathbf w$,
a near-optimal $\alpha(t)$,
%
then a good candidate is
%
\begin{equation}
\alpha(\tau) = \frac{1}{\lambda \, (\tau + t_0) },
\label{eq:alpha_trial}
\end{equation}
%
and to find the optimal value of $\lambda$.
%
Using Eq. \eqref{eq:alpha_trial}
in Eq. \eqref{eq:error_mbin1} yields
%
\begin{align}
\mathcal E
&=
\frac{1}{t+t_0}
\sum_{i=2}^n \frac{ \lambda_i^2 } { \lambda \, (2 \lambda_i - \lambda) }
\left[
  1 - \left(
    1 + \frac{t}{t_0}
  \right)^{1 - 2 \, \lambda_i/\lambda}
\right]
\notag
\\
&\stackrel{      t \gg t_0      }
          { =\joinrel=\joinrel= }
%
\sum_{i = 2}^n
\begin{dcases}
  \frac{ 1 }
       { t }
  \frac{ \lambda_i^2 }
       { \lambda \, (2 \, \lambda_i - \lambda) }
  &
  \mathrm{if \;} \lambda < 2 \, \lambda_i,
  \\%[1em]
  %
  %
  \frac{   1    }
       { 4 \, t }
  \ln \frac{ t + t_0 }
           {  t_0    }
  &
  \mathrm{if \;} \lambda = 2 \, \lambda_i,
  \\%[1em]
  %
  %
  \frac{ t_0^{ 2 \, \lambda_i / \lambda  - 1} }
       { t^{ 2 \, \lambda_i / \lambda } }
  \frac{ \lambda_i^2 }
       { \lambda \, (\lambda - 2 \, \lambda_i) }
  &
  \mathrm{if \;} \lambda > 2 \, \lambda_i.
\end{dcases}
\label{eq:y2_sum_perfect_invt}
\end{align}
%
\note{In the above expression,
$t + t_0$ are replaced by $t$.
%
Adding back $t_0$, we get
$$
\begin{aligned}
  \mathcal E
  \stackrel{         t \to \infty         }
           { =\joinrel=\joinrel=\joinrel= }
  \sum_{i = 2}^n
  \begin{dcases}
    \frac{ 1 }{ t + t_0 }
    \frac{ \lambda_i^2 } { \lambda \, (2 \, \lambda_i - \lambda) }
    &
    \mathrm{if \;} \lambda < 2 \, \lambda_i,
    \\%[1em]
    %
    %
    \frac{1}{4 \, (t + t_0) } \ln\frac{t + t_0}{t_0}
    &
    \mathrm{if \;} \lambda = 2 \, \lambda_i,
    \\
    %
    %
    \frac{ t_0^{ 2 \, \lambda_i / \lambda  - 1} }
         { (t + t_0)^{ 2 \, \lambda_i / \lambda } }
    \frac{ \lambda_i^2 }
         { \lambda \, (\lambda - 2 \, \lambda_i) }
    &
    \mathrm{if \;} \lambda > 2 \, \lambda_i.
  \end{dcases}
\end{aligned}
$$
}
%
The last two cases, $\lambda \ge 2\lambda_i$,
are asymptotically suboptimal
as they decay slower than $1/t$.
%
Thus, the optimal $\lambda$
satisfies $\lambda < 2 \min \lambda_i$,
and in this case,
we can seek that
the optimal value from
the smallest real root of
%
\begin{equation}
\sum_{i = 2}^n
\frac{ \lambda_i - \lambda }
{ \left(2 - \lambda/ \lambda_i \right)^2 }
= 0.
\label{eq:optimal_lambda_approx}
\end{equation}

Under the approximation, we have
$$
\frac{ \lambda_i^2 }{ \lambda \, (2 \lambda_i - \lambda) } \ge 1,
$$
with the equality achieved only at $\lambda = \lambda_i$,
we see the from Eq. \eqref{eq:y2_sum_perfect_invt}
that the minimal error is achieved with
$\lambda_2 = \dots = \lambda_n = \lambda$,
agreeing with the general conclusion
in Sec. \ref{sec:optWL}.

@@

As an example, consider the following matrix $\mathbf w$
%
\begin{equation}
  w_{ij} = \mu_{|i-j|} + \mu_{i+j-1} + \mu_{2n+1-i-j},
  \label{eq:w_band}
\end{equation}
%
where $\mu_0, \dots, \mu_b \, (b < n)$
are a set of numbers satisfying
%
\begin{equation}
\mu_0 + 2 \, \mu_1 + \cdots + 2 \, \mu_b = 1,
\label{eq:m_normalization}
\end{equation}
%
and $\mu_l = 0$ for $l > b$.

This matrix, being symmetric,
satisfies Eqs. \eqref{eq:w_sumj}-\eqref{eq:w_balance},
for the flat distribution $p_i \equiv 1/n$.
%
It has $n$ orthonormal eigenvectors,
$\pmb\varphi^{(1)}, \dots, \pmb\varphi^{(n)}$,
given by
%
\begin{equation}
\varphi^{(k)}_i
= \sqrt{
    \frac{ 2 - \delta_{k, 1} }
         {       n           }
       }
  \cos \frac{ \left(i - \frac12\right) (k - 1) \, \pi}{n}.
\label{eq:wband_eigenvector}
\end{equation}
%
with eigenvalues
\begin{equation}
  \lambda_k = \mu_0 + \sum_{l = 1}^b 2 \, \mu_l \cos \frac{(k - 1)  \, l \pi}{n}.
\label{eq:wband_eigenvalue}
\end{equation}
%
\note{We derive Eqs.
  \eqref{eq:wband_eigenvector} and \eqref{eq:wband_eigenvalue}
  below.
